{memes}: an R interface to the MEME Suite

### 2020-09-20 manuscript.Rmd
Removed old draft text, added more sections.

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing,comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

@@

The MEME Suite family of tools provides a
comprehensive set of utilities to interrogate motif content across a broad range
of biological contexts. However, several factors limit the full potential of
these tools for use in data analysis. 
  1) must do significant preprocessing to get data in correct input format
  2) Easily parseable output data often doesn't contain all information
    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
      - matrix is present in XML file, but no tools exist to parse this
  3) limited data visualization capabilities

@@

### Datatypes

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data. Motif
inputs are passed as `universalmotif` objects, another R/Bioconductor package
for representing motif matrices and their associated metadata. `memes` will also
always accept paths to files on disk, such as fasta files for sequence inputs,
and meme format files for motif inputs.

The MEME Suite tools do not return consistent data structures across tools. For
example, the tool TomTom, which compares query motifs to a database of known
motifs, returns tab-separated results which do not contain the frequency matrix
representations of the matched motifs. Instead, users must parse these matrices
back out from the input databases, creating additional friction for creating
novel data visualizations and exploratory analyses. Often, these data are
contained in XML formatted outputs, which although they hold all relevant
information, are difficult to parse. To our knowledge, no publicly available
tools exist to parse MEME Suite XML formatted data, therefore users must write
custom code to extract these data. `memes` exports data import functions for
each supported MEME Suite tool, which import these data as R data.frames. These
functions work internally to create the seamless file I/O when calling MEME
Suite tools, but the import functions can be called directly by users to import
previously generated data for use in R.

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query regions along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, and contain complex heirarchies. `memes` therefore extends the
`universalmotif` format by providing data.frames holding motif metadata which
retain links to their `universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames, and are
the basis for a majority of `memes` data outputs. These structures are also
valid input types to `memes` functions, and when used as such, output data are
appended as new columns to the input data, ensuring data provenance.

@@

`universalmotif` data.frames can additionally be used to examine or manipulate
the properties of any motif database. Using `universalmotif` functions to import
motif data, `as_universalmotif_dataframe()` converts any `universalmotif` object
into a `universalmotif` data.frame which can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames. Manipulation of linked
metadata columns will result in updating the linked `universalmotif` object when
calling `update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame into `universalmotif` format. In
this way, `memes` enables large-scale motif database manipulation & analysis
within the preexisting R framework.

### Genomic range-based data as entrypoint

**CONTINUE**

MEME Suite focuses on sequence data as the centerpoint of analysis, but in the
case of ChIP-seq, data are often stored as genomic coordinates with additional
metadata columns, such as `GRanges` objects in R/Bioconductor.

A key difference between `memes` and the MEME Suite tools is that `memes`
provides tools allowing the peak-level metadata to frame subsequent analysis.
For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
object containing the DNA sequence of each feature in the input ranges. If
ranges are split into groups by a metadata value, `get_sequence()` will return a
list of sequences for each group. When a sequence list is used as input into
`memes` core functions, the function will run once for each group, enabling a
native "split-apply-combine" workflow without any additional user overhead
\cite{splitapplycombine, wickham}. Together, these features empower users to
use range-based metadata as the centerpoint of their analyses.

### Old Design & Implementation
The key underlying design principle of `memes` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memes` tools can be used to extend
existing analysis pipelines, or to create novel analyses.

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files which together compose the full
results of the tool. While all tools produce machine-readable data formats, few
options exist for parsing the full complement of these data into human readable
format for use outside of the MEME Suite environment.

MEME Suite are run on the commandline and take 
as input fasta files of sequences
which will be searched for motifs, and if necessary, .meme format motif files
containing motifs to use in the search (REF: table of inputs).

At its most basic form, `memes` acts as a wrapper of these functions: users can
point `memes` tools to files on disk, `memes` will construct the commandline
call, output files will be saved on disk and imported into R for further analysis.
However, `memes` is more than just a simple wrapper of the MEME Suite, it is an
analysis framework allowing seamless integration of MEME Suite tools by
providing an abstraction layer which handles file input/output (I/O) to allow
inputs and outputs in the form of in-memory R objects.

When running a MEME Suite tool using `memes`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memes` tools can
be used as inputs to other `memes` tools, allowing easy construction of
pipelines (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).
 - unlike current MEME tools (ex tomtom -> FIMO is hard to do)

By shifting the focus from managing files on disk which may change infrequently to R data
objects, this framework reinforces the requirement of writing reproducible
analysis code to generate reproducible results.

Whenever possible, `memes` uses standard R data types for better integration
with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
way, users don't need to learn new ways of working with data, and outputs can
be integrated with other tools with limited effort.

Furthermore, `universalmotif` conversion tools allow `memes` detected motifs to
be used in many other popular Bioconductor motif packages, such as `TFBSTools`
and `motifStack`, providing the user with maximum flexibility in their analysis.

### Discriminative analysis
A powerful feature of several MEME Suite Tools is the ability to perform
discriminative analysis, where motifs are identified or tested for enrichment
in one group of sequences relative to another. In many instances, running tools
in discriminative mode on a set of control sequences outperforms simply using
shuffled input sequences as the control (back this up, change wording? maybe:
using biologically relevant control sequences can improve quality of results?).
Therefore, enabling complex discriminative analysis designs was central to the
design of `memes`.

When using list input to `memes` tools, names of list object members can be
passed to the `control` argument. This results in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences are combined
together to create a merged background set. In this way, performing data-driven
discriminative analysis is just as simple as one using shuffled sequences as
background.

```{r, eval=F}
############
# Replace this for a real set of peaks 
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")
############

# run Dreme on Static, Increasing, Decreasing sites using shuffled input as background
runDreme(sequences_list, "shuffle")

# run Dreme on Increasing, Decreasing sites using Static sites as background
runDreme(sequences_list, "Static")

# run Dreme on Decreasing sites using a combined set of Static + Increasing sites as background
runDreme(sequences_list, c("Static", "Increasing"))
```

## Metadata manipulation
`memes` is built on top of the `universalmotif` framework, which provides
methods for importing motifs from various file formats into a unified R data
representation complete with metadata information \cite{universalmotif}.
However, a major limitation of the `universalmotif` framework is that 
manipulating or filtering multiple `universalmotif` entries requires potentially
unfamiliar overhead for less experienced users, acting as a barrier to
performing potentially useful operations.

The `as_universalmotif_dataframe()` function from `memes` converts
`universalmotif` objects into a more familiar `data.frame` format which can be
manipulated using `tidyverse` or base R tools. These data.frames also contain a
special `motif` column which stores the `universalmotif` object of each entry.
Users can update columns in the data.frame which will be reflected in the
corresponding metadata entries of the `motif` by running `update_motifs()` after
manipulating the columns. These data.frames can be converted back into
`universalmotif` format using `as_universalmotif`, which also updates the
metadata using column values. The results from running the core MEME Suite
functions are stored in this format whenever appropriate, creating a consistent
framework for users to explore and manipulate motifs and their associated data values.

**This section is maybe too listy and could be fleshed out more / integrated above better**
Examples of common manipulations to motif databases which are aided by this
format include bulk renaming of entries, identification and removal of redundant
entries, and exploratory analysis of motif databases.

```{r}
# Example of renaming motifs w/ universalmotif syntax
library(universalmotif)
m1 <- create_motif()
m2 <- create_motif()
my_motifs <- list(m1,m2)

my_names <- c("motif one", "motif two")

for (i in 1:length(my_motifs)) {
  my_motifs[[i]]@name <- my_names[i]
}

# Example of renaming motifs w/ memes syntax
my_motifs <- list(m1,m2)
my_motifs %<>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = my_names) %>% 
  as_universalmotif()
```

This data format also allows users to integrate external data sources into their
analyses, for example, to filter a motif database to include only motifs for
genes which are expressed in their tissue of interest. Using motif databases
filtered in this way has been demonstrated in multiple contexts to provide
higher quality results when searching for novel regulators based on motif
content \cite{hector franco, eric olson, john stamm?, shep & greenleaf, klu worm
paper}.

## Data Visualization (section unfinished)

The MEME Suite provides a limited set of data visualizations which have limited
customizability. `memes` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

  - `view_tomtom_hits()` allows examination of tomtom results, allowing the user
  to evaluate the quality of the assigned match.

```{r}
# Example comparison of view_tomtom_hits results
plot_topHits <- example_tomtom %>% 
  dplyr::mutate(name = "Example Motif (Has Matches)") %>% 
  update_motifs() %>% 
  view_tomtom_hits(2)

plot_noMatch <- create_motif() %>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = "Example Motif (No Matches)") %>% 
  update_motifs() %>% 
  {
    x <- .
    x$tomtom <- list(NULL)
    return(x)
  } %>% 
  view_tomtom_hits()

cowplot::plot_grid(plotlist = list(plot_topHits[[1]], plot_noMatch[[1]]), ncol = 2, labels = "AUTO")
```

  - `ame_plot_heatmap()` allows comparison of AME results between multiple
  groups. Heirarchical ordering of hits shows commonalities and differences in enrichment between groups.
```{r}
# AME Heatmap example plots
example_ame$Decreasing %>% 
  ame_plot_heatmap()

example_ame %>% 
  dplyr::bind_rows(.id = "id") %>% 
  ame_plot_heatmap(group = id)
```

# Conclusion
 - `memes` is cool
 - use `memes`
 - it's on github (provide link)
 - manual is here:

# Notes

**Example of commandline vs `memes`?**
For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

The `memes` implementation of the above command would be written as follows:
```{r, eval=F}
# memes command setting random seed to 100
# user explicitly states they want to shuffle input sequences as control
runDreme("input.fasta", control = "shuffle", s = 100)
```

### Motif Comparison (TomTom)
 - takes runDreme/meme output as input -> pipes
 - enables using best matched motif as input to other tools (unlike TomTom on commandline)

### Motif Matching (FIMO)
 - returns locations as GRanges
 - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa

\begin{figure}
\includegraphics[width=\textwidth]{figures/memes_model.png}
\caption{A diagram of memes tools & workflows?
flowchart of each step, getsequence-> run-> import -> visualize -> goto run

Diagram how meme-suite tools work vs memes?
}
\label{fig:model} 
\end{figure}

# Future developments:
 - Additional meme suite tools
 - better support for PSP generation
 - Better support for amino-acid sequence usage (please contact us with feature requests)

>>>>

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

@@

The MEME Suite family of tools provides a
comprehensive set of utilities to interrogate motif content across a broad range
of biological contexts. However, several factors limit the full potential of
these tools for use in data analysis. 
  1) must do significant preprocessing to get data in correct input format
    - written to analyze 1 group of sequences, not amenable to multigroup designs
    - no common patterns for generating inputs from ChIP-seq whose data are primarily range based
  2) Easily parseable output data often doesn't contain all information
    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
      - matrix is present in XML file, but no tools exist to parse this
  3) limited data visualization capabilities
  4) difficult to integrate data with additional sources

@@

### Data Structures

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files containing varying data types.
`memes` reduces each MEME Suite tool to an R function which internally handles
file I/O such that all MEME Suite calling functions accept R objects as input,
and return R objects as output. Sequence inputs are stored in `BioStrings`
format, an R/Bioconductor package for storing biopolymer sequence data
\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
R/Bioconductor package for representing motif matrices and their associated
metadata \cite{universalmotif}. `memes` will also always accept paths to files
on disk, such as fasta files for sequence inputs, and meme format files for
motif inputs, reducing the need to read large files into memory to run MEME
Suite tools.

Although MEME Suite tools accept consistent data type inputs, they do not return
consistent data structures across tools. For example, the tool TomTom, which
compares query motifs to a database of known motifs, returns tab-separated
results which do not contain the frequency matrix representations of the matched
motifs. Instead, users must write custom code to parse these matrices back out
from the input databases, creating additional friction for creating novel data
visualizations and exploratory analyses. Often, output data are also returned in
XML format, which although they hold all relevant information, are difficult to
parse. To our knowledge, no publicly available tools exist to parse MEME Suite
XML formatted data, therefore users must write custom code to extract these
data. `memes` exports data import functions for each supported MEME Suite tool,
which import these data as R data.frames. These functions also work internally to
create the seamless file I/O when calling MEME Suite tools, but the import
functions can be called directly by users to import previously generated data
for use in R.

 - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
 - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
 - **Need to revise above statement, but note that most tools don't have support**

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex heirarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

@@

`universalmotif` data.frames are not limited to storing motifs from MEME Suite
data, and can additionally be used to examine or manipulate the properties of
any motif list or database. Using `universalmotif` functions to import motif data,
the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
into a `universalmotif` data.frame which can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames. Manipulation of linked
metadata columns will result in updating the linked `universalmotif` object when
calling `update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame back into `universalmotif` format. In
this way, `memes` enables large-scale motif database manipulation & analysis
within the preexisting R framework.

### Support for genomic range-based data

# RETURN
Motif analysis is important for assays like ChIP-seq which store data as genomic
coordinates, not sequence. This is in conflict with MEME Suite which is designed
to study sequences. Therefore, users must write custom code to extract DNA
sequence of their ranges of interest. Although tools like `bedtools getfasta`
can simplify the process of extracting DNA sequence from genomic coordinates,
some MEME tools require fasta headers to be specially formatted to extract full
functionality, requiring users to write additional custom code to properly
format input data.

`memes` provides the `get_sequence` function which automates the extraction of
DNA sequence from genomic coordinates while simultaneously producing MEME Suite
formatted fasta headers. `get_sequence` accepts genomic-range based inputs in
`GenomicRanges` format, the *de-facto* standard for genomic coordinate
representation in R \cite{genomicranges}. Other common genomic coordinate
representations, such as bed format, are easily imported as `GenomicRanges`
objects into R using preexisting import functions, meaning users must not write
any custom import functions to work with range-based data using `memes`. 
Sequences are returned in `Biostrings` format and can therefore be used as
inputs to all `memes` commands or as inputs to other R/Bioconductor functions
for sequence analysis.

During analysis of multiomic data, users often uncover different functional
categories of genomic regions through integration of multiple datasets. In these
instances, motif analysis acts as a powerful tool to examine whether differences
in motif content is associated with membership of distinct categories. In some
cases, performing differential motif analysis (using one category as the "test"
set and another as the "control") can uncover biologically relevant motifs which
are enriched in one category relative to another. Although the MEME Suite allows
differential enrichment testing, it does not inheritly provide a mechanism for
analyzing groups of sequences in parallel, or performing motif analysis with any
understanding of data categories. The `memes` framework enables "data-aware"
motif analysis workflows by allowing named lists of `Biostrings` objects as
input to each function. If using `GenomicRanges`, users can split peak data on a
metadata column using the base R `split` function, and use this result as input
to `get_sequence`, which will produce a list of `BioStrings` objects where each
entry is named after each data category from the split column. When a list is
used as input to a `memes` function, it runs the corresponding MEME Suite tool
for each object in the list. Users can also pass the name(s) of a category to
the `control` argument to automatically enable differential analysis of all
other list members against the control category sequences.  Therefore, `memes`
enables complex data-aware differential motif analysis workflows using simple
syntax to extend the capabilities of the MEME Suite.

### Table of supported MEME Suite Tools
 - comparison to other tools?
 - Maybe only do this if asked

## Data visualization
The MEME Suite provides a small set of data visualizations which have limited
customizability. `memes` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

 - `view_motifs()` plots motif as stack (if I decide to add this feature... need to think about this)
 - `ame_plot_heatmap` plot clustered heatmap of AME results 
   - `ame_compare_heatmap_methods` compare nonlinearity of data visualization in heatmap form
 - `view_tomtom_hits` visualize and compare TomTom matches to assess match quality
   - `force_best_match` can be used to update "best" match demonstrating how data visualization can be useful for data curation

The AME tool searches for enrichment of motifs within groups of sequences
relative to a control set. The meaningful result from this tool is the
statistical parameter (for example, a p-value) associated with the significance
of motif enrichment. However, AME does not provide a mechanism for visualizing
these results. Additionally, AME is often used 
## RETURN

## Containerized analysis maximizes availability and facilitates reproducibility
`memes` relies on a locally installed version of the MEME Suite which is
accessed by the users R process. Although R is available on Windows, Mac, and
Linux operating systems, the MEME Suite is incompatible with Windows, limiting
its adoption by Windows users. Additional barriers also exist to installing a
local copy of the MEME Suite on compatible systems. The MEME Suite relies on
several system-level dependencies whose installation can be difficult for novice
users. Finally, some tools in the MEME Suite use python to generate shuffled
control sequences for analysis, which presents a reproducibility issue as the
random number generation algorithm changed from python2.7 to python3. The MEME
Suite will build and install on both python2.7 and python3 systems, therefore
without careful consideration of this behavior, the same code run on two systems
may not produce identical results, even if using the same major version of the
MEME Suite. In order to democratize access to the MEME Suite on unsupported
operating systems, and to facilitate reproducibile motif analysis using `memes`, 
we have also developed docker and singularity (**NOTE:** singularity is broken
for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
preinstalled version of the MEME Suite along with an R/Bioconductor analysis
environment including the most recent version of `memes` and its dependencies.

# Future developments:
 - Additional meme suite tools
 - more data visualizations
 - better support for PSP generation
 - Better support for amino-acid sequence usage (please contact us with feature requests)

# Acknowledgements
 - Sean Davis automated docker builds
 - Megan Justice, feedback, feature requests
